---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# usethis

[![Travis build status](https://travis-ci.org/r-lib/usethis.svg?branch=master)](https://travis-ci.org/r-lib/usethis)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/r-lib/usethis?branch=master&svg=true)](https://ci.appveyor.com/project/r-lib/usethis)
[![Coverage status](https://codecov.io/gh/r-lib/usethis/branch/master/graph/badge.svg)](https://codecov.io/github/r-lib/usethis?branch=master)
[![CRAN status](http://www.r-pkg.org/badges/version/usethis)](https://cran.r-project.org/package=usethis)
[![lifecycle](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://www.tidyverse.org/lifecycle/#stable)

The goal of usethis is to automate many common package and analysis setup tasks.

## Installation

Install the released version of usethis from CRAN:

```{r, eval = FALSE}
install.packages("usethis")
```

Or install the development version from GitHub with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("r-lib/usethis")
```

## Usage

Most `use_*()` functions operate on the *active project*. If you've just used usethis to create a new package or project, that will be the current project. Otherwise usethis tries to confirm that current working directory can be recognized as a project. Use `proj_get()` and `proj_set()` for manual intervention. Some functions have no strong connections to projects and will expect you to provide a path.

usethis is quite chatty, explaining what it's doing and assigning you tasks. `✔` indicates something usethis has done for you. `●` indicates that you'll need to do some work yourself.

Below is a quick look at how usethis can help to set up a package.

*Note: usethis is gaining more and more functionality for analytical projects that are not packages. Stay tuned.*

```{r}
library(usethis)

# Create a new package -------------------------------------------------
tmp <- file.path(tempdir(), "mypkg")
create_package(tmp)

# Modify the description ----------------------------------------------
use_mit_license("My Name")

use_package("MASS", "Suggests")

use_dev_package("callr")

# Set up various packages ---------------------------------------------
use_roxygen_md()

use_rcpp()

use_revdep()

# Set up other files -------------------------------------------------
use_readme_md()

use_news_md()

use_test("my-test")

x <- 1
y <- 2
use_data(x, y)

# Use git ------------------------------------------------------------
use_git()
```

Please note that the usethis project is released with a [Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.
