---
title: "Exploring badge accessibility"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(usethis)
```

Various functions in usethis insert a badge into a package's README file.
A badge consists of an image with information about the package, which is usually hyperlinked to a location with further details.
Here is a typical badge:

<!-- badges: start -->
[![badge_example](https://img.shields.io/badge/badge%20label-badge%20message-yellowgreen)](https://shields.io/)
<!-- badges: end -->

For example, the badge message might be "passing" or "failing", reflecting the `R CMD check` status of its current development state, and then link to the actual log file.
Or the badge could display the current test coverage, as a percentage of lines, and then link to a site where one can explore test coverage line-by-line.

<!-- badges: start -->
[![R-CMD-check](https://github.com/r-lib/usethis/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/r-lib/usethis/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/r-lib/usethis/branch/main/graph/badge.svg)](https://app.codecov.io/gh/r-lib/usethis?branch=main)
<!-- badges: end -->

A key point is that the badge often presents *dynamic* information.

We'd like to improve the user experience for those consuming the README using assistive technology, such as a screen reader.

There are at least two ways that usethis can influence this:

* The construction of the hyperlinked badge. For example, we control the alt
  text.
* The image itself. We can encourage the badge providers to build accessibility
  into their badge SVGs. If the "native" badge is not accessible, usethis can
  substitute an accessible badge from a 3rd party, such as
  [shields.io](https://shields.io/).

This article is meant to gather badge facts in one place and facilitate a dialogue with those using assistive technology.

## Markdown image link

usethis currently inserts badges in a special "badge block" in the README, fenced with HTML comment tags.
Within the badge block, each badge is a hyperlinked image, with alt text.
Here is the basic form:

```
[![alt_text](http://path/to/the/badge/image)](http://path/to/more/details)
```

Note that the `alt_text` should be a static label explaining the badge's purpose, whereas the dynamic information is baked into the externally-served image.

usethis definitely controls this `alt_text`, so that is something I'd like feedback on.

## Badge image

Most badges these days tend to be SVG files, which is fortunate, since SVGs can carry attributes to support accessibility.
We demonstrate this with a custom badge generated via the [shields.io](https://shields.io/) web service.

I will request a badge with this URL:

```
https://img.shields.io/badge/my__label-my__message-orange
```

Here it is as a badge, with the hyperlink set to <https://shields.io/>.
The badge is inside an HTML comment block, like usethis does for README badges.
I've prefixed every field I can influence with "my"; there are 3 such fields:

1. The alt text.
2. The label. This static text appears on the left side of the badge, on a grey
   background, and describes the badge.
3. The message. This text appears on the right side of the badge, on a colored
   background. This is often the dynamic part, where both the color and the text
   convey a dynamic fact, e.g. "passing" on a green background or "failing" on a
   red background.

<!-- badges: start -->
[![my_alt_text](https://img.shields.io/badge/my__label-my__message-orange)](https://shields.io/)
<!-- badges: end -->

Now we inspect the badge SVG.
I apologize in advance for the amount of data displayed here, but it seems good to show one badge in full gory detail.
Later, we will display badge information very selectively.

```{r R.options = list(tidyverse.quiet = TRUE)}
library(xml2)
library(purrr)

inspect_badge <- function(badge) {
  badge |>
    read_xml() |>
    as_list() |>
    pluck("svg")
}

inspect_badge("https://img.shields.io/badge/my__label-my__message-orange")
```

It is my understanding that the two main pieces of metadata are the `title` and the `aria-label` attribute.
Here I reveal just those two items from the badge:

```{r}
inspect_badge <- function(badge) {
  x <- badge |>
    read_xml() |>
    as_list() |>
    pluck("svg")
  list(
    title = pluck(x, "title", 1),
    `aria-label` = pluck(x, attr_getter("aria-label"))
  )
}

inspect_badge("https://img.shields.io/badge/my__label-my__message-orange")
```

This badge carries the same information in the `title` and in `aria-label`, which is "my_label: my_message".
I would be interested to learn more about why the same information is included twice and if that is a good or bad thing for the screen reader experience.

### shields.io badges are accessible today

One of the reasons I inspected a shields.io badge is that this may provide a usable alternative for any service whose official badge is not (yet?) screen-reader-friendly.

The custom badge above is completely static.
But shields.io also supports custom dynamic badges, when the necessary information (label, message, color) is available from a suitable JSON endpoint.
Finally, and most relevant to usethis, shields.io offers pre-configured dynamic badges for the most commonly "badged" services, including GitHub Actions and Codecov.

Here is a shields.io badge for usethis's `R CMD check` workflow on GitHub Actions:

<!-- badges: start -->
[![my_alt_text-R-CMD-check](https://img.shields.io/github/workflow/status/r-lib/usethis/R-CMD-check?label=my_label-R-CMD-check)](https://github.com/r-lib/usethis/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Again, I am indicating fields I control with `my_alt_text` and `my_label`, so it's easier to get feedback on what usethis should do.

Here is the `title` and `aria-label` of the badge above:

```{r}
inspect_badge("https://img.shields.io/github/workflow/status/r-lib/usethis/R-CMD-check?label=my_label-R-CMD-check")
```

Now we will take inventory of the main badges inserted by usethis and where things stand re: accessibility.

## `R CMD check` status

We generally obtain the current `R CMD check` status from a GitHub Actions workflow.

Here is the official badge provided by GitHub:

<!-- badges: start -->
[![R-CMD-check](https://github.com/r-lib/usethis/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/r-lib/usethis/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Here is the `title` and `aria-label` of the badge above:

```{r}
inspect_badge("https://github.com/r-lib/usethis/actions/workflows/R-CMD-check.yaml/badge.svg")
```

At the time of writing (late December 2021), the badge does not include such information.

I have requested this in GitHub's Actions and Packages Feedback forum and the response from GitHub is encouraging.
Hopefully the native badge will gain improved accessibility early in 2022.

<https://github.com/github/feedback/discussions/8974>

In the meantime, one could use a shields.io badge to report `R CMD check` status, as demonstrated in the previous section.
A maintainer could do this as a one-off or, if the GitHub badge upgrade is slow in coming, we could implement it in `usethis::use_github_actions_badge()`.

## Package lifecycle

`usethis::use_lifecycle_badge()` declares the developmental stage of a package, using the framework from <https://lifecycle.r-lib.org/articles/stages.html>.
This function already inserts a shields.io badge:

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
<!-- badges: end -->

Here is the `title` and `aria-label` of the badge above:

```{r}
inspect_badge("https://img.shields.io/badge/lifecycle-stable-brightgreen.svg")
```

It's possible that usethis should be using a badge "owned" by the lifecycle package.
I think we don't do so now, because declaring the lifecycle stage of a whole package does not necessarily imply the need to take a formal dependency on the lifecycle package, which is usually what causes badges to be copied into the `man/figures/` directory.

Also, the badges shipped by the lifecycle package are _not_ currently accessible, which is another reason not to use them.
But I have opened an issue about this (<https://github.com/r-lib/lifecycle/issues/117>).
This should be a relatively easy fix, since these are static badges.
Once done, package maintainers would need to update the SVGs that are stored in `man/figures/`.

## CRAN status

`usethis::use_cran_badge()` places a badge that indicates what package version is available on CRAN.
This badge is served by METACRAN (<https://www.r-pkg.org>) and maintainer Gábor Csárdi has already incorporated `aria-label` into the badges (in [this commit](https://github.com/metacran/metacranweb/commit/8287a21e6dc2bc50a2d8a7b5a5a56904ae1d04ff)).
All available badges are listed [here](https://www.r-pkg.org/services).

Here's the badge placed by `usethis::use_cran_badge()`:

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/usethis)](https://CRAN.R-project.org/package=usethis)
<!-- badges: end -->

Here is the `title` and `aria-label` of the badge above:

```{r}
inspect_badge("https://www.r-pkg.org/badges/version/usethis")
```

At the time of writing (late December 2021), `aria-label` is present, but `title` is not.
I would be interested to know if this is a good situation for those using a screen reader.

## Code coverage

`usethis::use_coverage(type = c("codecov", "coveralls", ...)` calls the internal helper `usethis:::use_codecov_badge()` to insert a badge for either the Codecov or Coveralls service.

Here are examples of those two badges (note that usethis does not use Coveralls, so I'm using a different package to demo):

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/r-lib/usethis/branch/main/graph/badge.svg)](https://app.codecov.io/gh/r-lib/usethis?branch=main)
[![Coveralls test coverage](https://coveralls.io/repos/github/trinker/sentimentr/badge.svg?branch=master)](https://coveralls.io/github/trinker/sentimentr?branch=master)
<!-- badges: end -->

Here are the `title` and `aria-label` of those badges::

```{r}
inspect_badge("https://codecov.io/gh/r-lib/usethis/branch/main/graph/badge.svg")

inspect_badge("https://coveralls.io/repos/github/trinker/sentimentr/badge.svg?branch=master")
```

At the time of writing (late December 2021), neither badge offers a `title` or `aria-label`.

Here are badges from shields.io for Codecov and Coveralls:

<!-- badges: start -->
[![Codecov test coverage](https://img.shields.io/codecov/c/github/r-lib/usethis?label=test%20coverage&logo=codecov)](https://app.codecov.io/gh/r-lib/usethis?branch=main)
[![Coveralls test coverage](https://coveralls.io/repos/github/trinker/sentimentr/badge.svg?branch=master)](https://img.shields.io/coveralls/github/trinker/sentimentr?logo=coveralls)
<!-- badges: end -->

Here are the `title` and `aria-label` of the shields.io badges::

```{r}
inspect_badge("https://img.shields.io/codecov/c/github/r-lib/usethis?label=test%20coverage&logo=codecov")

inspect_badge("https://img.shields.io/coveralls/github/trinker/sentimentr?logo=coveralls")
```

## Bioconductor

If a package is on Bioconductor, `usethis::use_bioc_badge()` can be used to insert a badge for its Bioconductor build status.

Here is such a badge:

<!-- badges: start -->
[![BioC status](http://www.bioconductor.org/shields/build/release/bioc/biocthis.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/biocthis)
<!-- badges: end -->

Here is the `title` and `aria-label` of that badge:

```{r}
inspect_badge("http://www.bioconductor.org/shields/build/release/bioc/biocthis.svg")
```

At the time of writing (late December 2021), the badge does not have a `title` or `aria-label`.

It's not immediately clear how to resolve this, but I suspect that Bioconductor would be receptive to a request to create more accessible badges.
