---
title: "Converting usethis's UI to use cli"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  out.width = "100%",
  #asciicast_knitr_output = "html",
  asciicast_theme = "solarized-light", # tango, solarized-dark, solarized-light
  asciicast_cols = 72
)
asciicast::init_knitr_engine(
  #echo = TRUE,
  #echo_input = FALSE,
  startup = quote({
    library(usethis)
    library(glue)
    options(cli.num_colors = 256)
    ui_code_snippet <- usethis:::ui_code_snippet
    usethis_theme <- usethis:::usethis_theme
    ui_inform <- usethis:::ui_inform
    kv_line <- usethis:::kv_line
    ui_special <- usethis:::ui_special
    ui_abort <- usethis:::ui_abort
    ui_bullets <- usethis:::ui_bullets
    set.seed(1) })
)
```

```{r setup}
library(usethis)
library(glue)
```

*In a hidden chunk here, I'm "exporting" some unexported internal helpers, so that I can use them and talk about them. For similar reasons, I attach glue above, so that certain glue functions work here, without explicitly namespacing them.*

```{r}
#| include: false
ui_code_snippet <- usethis:::ui_code_snippet
usethis_theme <- usethis:::usethis_theme
ui_inform <- usethis:::ui_inform
kv_line <- usethis:::kv_line
ui_special <- usethis:::ui_special
ui_abort <- usethis:::ui_abort
ui_bullets <- usethis:::ui_bullets
```

## Block styles

The block styles exist to produce bulleted output with a specific symbol, using a specific color.

```{asciicast}
#| collapse: false
f <- function() {
  ui_todo("ui_todo(): red bullet")
  ui_done("ui_done(): green check")
  ui_oops("ui_oops(): red x")
  ui_info("ui_info(): yellow i")
  ui_line("ui_line(): (no symbol)")
}
f()
```

Another important feature is that all of this output can be turned off package-wide via the `usethis.quiet` option.

```{asciicast}
withr::with_options(
  list(usethis.quiet = TRUE),
  ui_info("You won't see this message.")
)
withr::with_options(
  list(usethis.quiet = FALSE), # this is the default
  ui_info("But you will see this one.")
)
```

These styles are very close to what can be done with `cli::cli_bullets()` and the way it responds to the names of its input `text`.

```{asciicast}
cli::cli_bullets(c(
        "noindent",
  " " = "indent",
  "*" = "bullet",
  ">" = "arrow",
  "v" = "success",
  "x" = "danger",
  "!" = "warning",
  "i" = "info"
))
```

A direct translation would look something like this:

| Legacy `ui_*()` | `cli_bullets()` shortcode | tweaks needed                 |
|-----------------|---------------------------|-------------------------------|
| `ui_todo()`     | `*`                       | blue (default) -\> red        |
| `ui_done()`     | `v`                       | perfect match                 |
| `ui_oops()`     | `x`                       | perfect match                 |
| `ui_info()`     | `i`                       | blue (default) -\> yellow     |
| `ui_line()`     | (unnamed)                 | sort of a perfect match? although sometimes `ui_line()` is used just to get a blank line |

The overall conversion plan is to switch to a new function, `ui_bullets()`, which is a wrapper around `cli::cli_bullets()`, that adds a few features:

* Early exit, without emitting messages, if the `usethis.quiet` option is `TRUE`.
* A usethis theme, that changes the color of certain bullet styles and adds a new style for todo's.

```{asciicast}
ui_bullets(c(
  "v" = "A great success!",
  "_" = "Something you need to do.",
  "x" = "Bad news.",
  "i" = "The more you know.",
  " " = "I'm just here for the indentation.",
  "No indentation at all. Not used much in usethis."
))
```


Summary of what I've done for todo's:

-   Introduce a new bullet shortcode for a todo. Proposal: `_` (instead of `*`), which seems to be the best single ascii character that evokes a place to check something off.
-   Use `cli::symbol$checkbox_off` as the symbol (instead of a generic bullet). I guess it will continue to be red.

In terms of the block styles, that just leaves `ui_code_block()`, which is pretty different.
`ui_code_block()` is used to put some code on the screen and optionally place it on the clipboard.
I have created a new function, `ui_code_snippet()` that is built around `cli::code_block()`.
Main observations:

* `cli::code_block(language = "R")` applies syntax highlighting and hyperlinks (e.g. to function help topics) to R code, which is cool. Therefore the `language` argument is also exposed in `ui_code_snippet()`, defaulting to `"R"`. Use `""` for anything that's not R code:
    ```{asciicast}
    ui_code_snippet("x <- 1 + 2")
    ui_code_snippet("#include <blah.h>", language = "")
    ```
* `ui_code_snippet()` takes a scalar glue-type template string or a vector of lines. Note that the two calls below produce the same output.
    ```{asciicast}
    ui_code_snippet("
      options(
        warnPartialMatchArgs = TRUE,
        warnPartialMatchDollar = TRUE,
        warnPartialMatchAttr = TRUE
      )")
    # produces same result as
    ui_code_snippet(c(
      "options(",
      "  warnPartialMatchArgs = TRUE,",
      "  warnPartialMatchDollar = TRUE,",
      "  warnPartialMatchAttr = TRUE",
      ")"))
    ```
* `ui_code_snippet()` does glue interpolation, by default, before calling `cli::cli_code()`, which means you have to think about your use of `{` and `}`. If you want literal `{` or `}`:
  - Use `interpolate = FALSE`, if you don't need interpolation.
  - Do the usual glue thing and double them, i.e. `{{` or `}}`.
  - If this becomes a real pain, open an issue/PR about adding `.open` and `.close` as arguments to `ui_code_snippet()`.

## Utility functions

The block style functions all route through some unexported utility functions.

`is_quiet()` just consults the `usethis.quiet` option and implements the default of `FALSE`.

```{r}
#| eval: false
is_quiet <- function() {
  isTRUE(getOption("usethis.quiet", default = FALSE))
}
```

`ui_bullet()` is an intermediate helper used by `ui_todo()`, `ui_done()`, `ui_oops()` and `ui_info()`.
It does some hygiene related to indentation (using the `indent()` utility function), then calls `ui_inform()`.
`ui_line()` and `ui_code()` both call `ui_inform()` directly.

`ui_inform()` is just a wrapper around `rlang::inform()` that is guarded by a call to `is_quiet()`

```{r}
#| eval: false
ui_inform <- function(...) {
  if (!is_quiet()) {
    inform(paste0(...))
  }
  invisible()
}
```

Other than `is_quiet()`, which will continue to play the same role, I anticipate that we no longer need these utilities (`indent()`, `ui_bullet()`, `ui_inform()`).
Updates from the future:

* `indent()` turns out to still be useful in `ui_code_snippet()`, so I've inlined
  it there, to avoid any reliance on definitions in ui-legacy.R.
* `ui_bullet()` has been renamed to `ui_legacy_bullet()` for auto-completion happiness with the new `ui_bullets()`.

Let's cover `ui_silence()` while we're here, which *is* exported.
It's just a `withr::with_*()` function for executing `code` with `usethis.quiet = TRUE`.

```{r}
#| eval: false
ui_silence <- function(code) {
  withr::with_options(list(usethis.quiet = TRUE), code)
}
```

## Inline styles

### Legacy functions

usethis has its own inline styles (mostly) for use inside functions like `ui_todo()`:

* `ui_field()`
* `ui_value()`
* `ui_path()`
* `ui_code()`
* `ui_unset()`

```{asciicast}
# why is this block truncated from the top in the rendered document?
new_val <- "oxnard"
x <- glue("{ui_field('name')} set to {ui_value(new_val)}")
dput(x)
ui_done(x)
```

The inline styles enact some combination of:

* Color, e.g. `crayon::green(x)`
* Collapsing a collection of things to one thing, e.g. `c("a", "b", "c")` to "a, b, c"
* Quoting, e.g. `encodeString(x, quote = "'")`

`ui_path()` is special because it potentially modifies the input before styling it.
`ui_path()` first makes the path relative to a specific base (by default, the active project root) and, if the path is a directory, it also ensures there is a trailing `/`.

`ui_unset()` is a special purpose helper used when we need to report that something is unknown, not configured, nonexistent, etc.

```{asciicast}
# why is this block truncated from the top in the rendered document?
x <- glue("Your super secret password is {ui_unset()}.")
dput(x)
ui_info(x)
```

### cli replacements

In general, we can move towards cli's native inline-markup:
<https://cli.r-lib.org/reference/inline-markup.html>

Here's the general conversion plan:

-   `ui_field()` becomes `{.field blah}`. In `usethis_theme()`, I tweak the `.field` style to apply single quotes if color is not available, which is what `ui_field()` has always done.
-   `ui_value()` becomes `{.val value}`.
-   `ui_path()` is connected to `{.path path/to/thing}`, but, as explained above, `ui_path()` also does more. Therefore, I abstracted the "path math" into an internal helper, `ui_path_impl()`, which is aliased to `pth()` for compactness. Here's a typical conversion:
    ```{r}
    #| eval: false
    # using legacy functions
    ui_done("Setting {ui_field('LazyData')} to \\
             {ui_value('true')} in {ui_path('DESCRIPTION')}")
    # using new cli-based ui
    ui_bullets(c(
      "v" = "Setting {.field LazyData} to {.val true} in {.path {pth('DESCRIPTION')}}."
    ))
    ```
    It would be nice to create a custom inline class, e.g. `{.ui_path {some_path}}`, which I have done in, e.g., googledrive.
    But it's not easy to do this *while still inheriting cli's `file:` hyperlink behaviour*, which is very desirable.
    So that leads to the somewhat clunky, verbose pattern above, but it gives a nice result.
-   `ui_code()` gets replaced by various inline styles, depending on what the actual goal is, such as:
    -   `{.code some_text}`
    -   `{.arg some_argument}`
    -   `{.cls some_class}`
    -   `{.fun devtools::build_readme}`
    -   `{.help some_function}`
    -   `{.run usethis::usethis_function()}`
    -   `{.topic some_topic}`
    -   `{.var some_variable}`
- `ui_unset()` is replaced by `ui_special()`, which you'll see more of below. Currently the intended grey color doesn't show up when I render this document using solarized-dark and so far I can't get to the bottom of that :( Why isn't it the same grey as "[Copied to clipboard]" in `ui_code_snippet()`, which does work?

## Conditions

I'm moving from `ui_stop()`:

```{r}
#| eval: false
ui_stop <- function(x, .envir = parent.frame()) {
  x <- glue_collapse(x, "\n")
  x <- glue(x, .envir = .envir)

  cnd <- structure(
    class = c("usethis_error", "error", "condition"),
    list(message = x)
  )

  stop(cnd)
}
```

to `ui_abort()`:

```{r}
#| eval: false
ui_abort <- function(message, ..., class = NULL, .envir = parent.frame()) {
  cli::cli_div(theme = usethis_theme())
  # bullet naming gymnastics, see below
  cli::cli_abort(
    message,
    class = c(class, "usethis_error"),
    .envir = .envir,
    ...
  )
}
```

The main point of `ui_abort()` is to use to `cli_abort()` (and to continue applying the `"usethis_error"` class).

I also use `ui_abort()` to apply different default bullet naming/styling.
Starting with `"x"` and then defaulting to `"i"` seems to fit best with usethis's existing errors.

```{asciicast}
#| error: true
# why is this block truncated from the top in the rendered document?
block_start = "# <<<"
block_end = "# >>>"
ui_abort(c(
  "Invalid block specification.",
  "Must start with {.code {block_start}} and end with {.code {block_end}}."
))
```

Any bullets that are explicitly given are honored.

```{asciicast}
#| error: true
ui_abort(c("v" = "It's weird to give a green check in an error, but whatever."))
ui_abort(c(
  "!" = "Things are not ideal.",
  ">" = "Look at me!"
))
```

`rlang::abort()` and `cli::cli_abort()` start with `"!"` by default, then use `"*"` and `" "`, respectively.

The legacy functions also include `ui_warn()`.
It has very little usage and, instead of converting it, I've eliminated its use altogether in favor of a `"!"` bullet:

```{asciicast}
ui_bullets(c("!" = "The guy she told you not to worry about."))
```

Sidebar: Now that I'm looking at a lot of the new errors with `ui_abort()` I realize that usethis also needs to be passing the `call` argument along.
I'm going to leave that for a future, separate effort.

## Sitrep and format helpers

This is a small clump of  functions that support sitrep-type output.

-   `hd_line()` *unexported and, apparently, unused! now removed*
-   `kv_line()` *unexported, so has new cli implementation*
-   `ui_unset()` *exported and succeeded by `ui_special()`*

`kv_line()` stands for "key-value line".
Here's what it used to be:

```{asciicast}
kv_line_legacy <- function(key, value, .envir = parent.frame()) {
  value <- if (is.null(value)) ui_unset() else ui_value(value)
  key <- glue(key, .envir = .envir)
  ui_inform(glue("{cli::symbol$bullet} {key}: {value}"))
}

url <- "https://github.com/r-lib/usethis.git"
remote <- "origin"
kv_line_legacy("URL for the {ui_value(remote)} remote", url)

host <- "github.com"
kv_line_legacy("Personal access token for {ui_value(host)}", NULL)
```

Key features:

* Interpolates data and applies inline style to the result.
  Works differently for `key` and `value`, because you're much more likely to use interpolation and styling in `key` than `value`.
* Has special handling when `value` is `NULL`.
* Applies `"*"` bullet name/style to over all result.

I won't show the updated source for `kv_line()` but here is some usage to show what it's capable of:

```{asciicast}
# why is this block truncated from the top in the rendered document?
noun <- "thingy"
value <- "VALUE"
kv_line("Let's reveal {.field {noun}}", "whatever")

kv_line("URL for the {.val {remote}} remote", I("{.url {url}}"))

kv_line("Personal access token for {.val {host}}", NULL)

kv_line("Personal access token for {.val {host}}", ui_special("discovered"))
```

`ui_special()` is the successor to `ui_unset()`.

## Questions

There's currently no drop-in substitute for `ui_yeah()` and `ui_nope()` in cli.
Related issues: <https://github.com/r-lib/cli/issues/228>, <https://github.com/r-lib/cli/issues/488>.
Therefore, in the meantime, `ui_yeah()` and `ui_nope()` are not-quite-superseded for external users.

However, internally, I've switched to the unexported functions `ui_yep()` and `ui_nah()` that are lightly modified versions of `ui_yeah()` and `ui_nope()` that use cli for styling.

```{r}
#| eval: false
if (ui_nope("
      Current branch ({ui_value(actual)}) is not repo's default \\
      branch ({ui_value(default_branch)}).{details}")) {
        ui_abort("Cancelling. Not on desired branch.")
    }
```

## Miscellaneous notes

* I've been adding a period to the end of messages, as a general rule.
* In terms of whitespace and indentation, I've settled on some conventions.
  The overall goal is to get the right user-facing output (obviously), while making it as easy as possible to _predict_ what that's going to look like when you're writing the code.
    ```{r}
    #| eval: false
    ui_bullets(c(
      "i" = "Downloading into {.path {pth(destdir)}}.",
      "_" = "Prefer a different location? Cancel, try again, and specify
             {.arg destdir}."
    ))
    ...
    ui_bullets(c("x" = "Things are very wrong."))
    ```
  Key points:
  - Put `ui_bullets(c(` on its own line, then all of the bullet items, followed by `))` on its own line. Sometimes I make an exception for a bullet list with exactly one, short bullet item.
  - Use hard line breaks inside bullet text to comply with surrounding line length. In subsequent lines, use indentation to get you just past the opening `"`. This extraneous white space is later rationalized by cli, which handles wrapping.
  - Surround bullet names like `x` and `i` with quotes, even though you don't have to, because it's required for other names, such as `!` or `_` and it's better to be consistent.
  - Here's another style I like that applies to `ui_abort()`, where there's just one, unnamed bullet, but the call doesn't fit on one line.
    ```{r}
    #| eval: false
    pr <- list(pr_number = 13)
    ui_abort("
      The repo or branch where PR #{pr$pr_number} originates seems to have been
      deleted.")
    ```

TODO: come back to `pr_push()` and deal with that menu stuff. Ditto for `choose_branch()` and `choose_pr()`. Ditto for `target_repo()` in utils-github.R.

TODO: come back to this bit in git.R: `purrr::walk(format(cfg), ui_legacy_bullet)`

TODO: NEWS bullet

TODO: Update principles.md
