---
title: "usethis setup"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

You will get the most out of usethis if you do some setup. These setup tasks do not need to be done all at once or even done at all. But usethis can offer the most support for package development and Git/GitHub workflows with some advance configuration. usethis can even help you with this!

Key steps that accelerate your R development workflow (details on *how* to do all this follow):

  * Make usethis available in interactive R sessions.
  * Provide usethis with info to use in all new R packages you create.
  * Use the "sitrep" functions to get a health check or gather info when you're
    going to ask for help.
  * Sign up for a free GitHub.com account, if you plan to use GitHub.
  * Install Git.
  * Configure your Git `user.name` and `user.email`.
  * If you use RStudio, make sure RStudio can find your Git executable. If you use GitHub, make sure you can pull/push from your local computer to GitHub.com, in general and from RStudio.
  * Get a personal access token from GitHub.com and make it available in R sessions.
  * Prepare your system to build R packages from source.
  
## Use usethis or devtools in interactive work

The usethis package was carved out of the [devtools package](https://devtools.r-lib.org) as part of its "conscious uncoupling" in [the v2.0.0 release](https://www.tidyverse.org/articles/2018/10/devtools-2-0-0/). But note that devtools makes all of usethis's functions available and they should feel like they are part of devtools itself. In addition, devtools offers some functions of its own and exposes selected functionality from a few other packages. You might enjoy making devtools (and therefore usethis) available in all your interactive R work.

Call `usethis::use_devtools()` for prompts to do this:

```{r, eval = FALSE}
library(devtools) ## or library(usethis)

use_devtools()
```

This will:

  * Open your `.Rprofile` startup file for editing.
  * Show the necessary code snippet in the console and put it on your clipboard.
  * Prompt you to paste, save, close, restart R. Done!

The suggested snippet looks like this:

```{r, eval = FALSE}
if (interactive()) {
  suppressMessages(require(devtools))
}
```

More resources on `.Rprofile`:

  * The official [Startup documentation](https://stat.ethz.ch/R-manual/R-patched/library/base/html/Startup.html)
  * The [Startup chapter](https://whattheyforgot.org/r-startup.html) of the What They Forgot to Teach You About R workshop materials
  * Kevin Ushey's blog post [RProfile Essentials](http://kevinushey.github.io/blog/2015/02/02/rprofile-essentials/)

## Store default values for DESCRIPTION fields and other preferences

Certain options are consulted by usethis and allow you to set personal defaults:

  * `usethis.full_name`: consulted for making, e.g., supplemental license files.
  * `usethis.protocol`: specifies your preferred transport protocol for
    Git. Either "https" or "ssh". See the help for `git_protocol()` for more.
  * `usethis.description`: named list of default DESCRIPTION fields for new
    packages made with `usethis::create_package()`.
  * `usethis.quiet`: if `TRUE`, prevents usethis from printing messages to the
    console.
  * `usethis.destdir`: a default directory to use in `create_from_github()` and
    `use_course()`.

Define any of these options in your `.Rprofile`, which can be opened for editing via `usethis::edit_r_profile()`. Here is example code:

```{r, eval = FALSE}
options(
  usethis.full_name = "Jane Doe",
  usethis.protocol  = "ssh",
  usethis.description = list(
    "Authors@R" = utils::person(
        "Jane", "Doe",
        email = "jane@example.com",
        role = c("aut", "cre"),
        comment = c(ORCID = "JANE'S-ORCID-ID")
    ),
    Version = "0.0.0.9000"
  ),
  usethis.destdir = "~/the/place/where/I/keep/my/R/projects"
)
```

Save similar code in your `.Rprofile` and restart R for it to take effect.

## The "sitrep" functions

These functions gather information that help you or others troubleshoot things:

  * `proj_sitrep()`: prints info about the active usethis project, working
    directory, and the active RStudio Project. Points out when things are
    peculiar and how to fix.
  * `git_sitrep()`: prints info about your current Git, gert, and GitHub setup.
  
"Sitrep" is short for "**sit**uation **rep**ort".

## Get a GitHub.com account

Sign up for a free account with [GitHub.com](https://github.com/). [Happy Git and GitHub for the useR](https://happygitwithr.com/github-acct.html) provides more advice about picking your username.

## Install Git

Please see [Happy Git and GitHub for the useR](https://happygitwithr.com/install-git.html) for instructions on how to install Git. It is beyond the scope of this article.

usethis itself does not actually need the Git that you install, because it uses the [gert](https://docs.ropensci.org/gert/) package which wraps [libgit2](https://libgit2.github.com). But, chances are, you want to do normal Git things, like diff and commit and push, from RStudio or in the shell and for that you must install Git.

## Configure `user.name` and `user.email`

Once Git is installed, introduce yourself to Git.

```{r, eval = FALSE}
library(usethis) ## or library(devtools)
use_git_config(user.name = "Jane Doe", user.email = "jane@example.com")

## check by running a git situation-report: 
##   - your user.name and user.email should appear under "User"
git_sitrep()
```

`usethis::use_git_config()` helps you configure your `user.name` and `user.email`. Substitute **your name** and **your email address**.

What user name should you give to Git? This does not have to be your GitHub username, although it can be. Another good option is your actual first name and last name. Your commits will be labelled with this name, so this should be informative to potential collaborators.

What email should you give to Git? This must be the email associated with your GitHub account.

`usethis::git_sitrep()` generates a git situation-report. It can help you confirm things will work as expected; it can also help you diagnose problems.

### Equivalent Git commands

The code chunk above is doing the equivalent of this:

```{sh, eval = FALSE}
git config --global user.name 'Jane Doe'
git config --global user.email 'jane@example.com'
git config --global --list
```

### Optional: configure Git's editor

Another Git option that many people eventually configure is the editor. This will come up if you use Git from a shell. At some point, you will fail to give Git what it wants in terms of a commit message and it will kick you into an editor. This can be distressing, if it’s not your editor of choice and you don’t even know how to save and quit. You can enforce your will by executing this in R:

```{r, eval = FALSE}
library(usethis)

use_git_config(core.editor = "nano")
```

To do the same thing with command line Git, execute this in a shell:

```{sh, eval = FALSE}
git config --global core.editor emacs
```

Substitute your preferred editor for emacs here. A popular choice is [nano](https://www.nano-editor.org/). The default, if you don't configure `core.editor`, is usually vim.

## Connections: Git, GitHub, RStudio

As stated above, usethis doesn't actually use the Git you install and has no absolute requirement that you use GitHub or use RStudio. But use of usethis is highly correlated with the desire to do all of these things, in a pleasant way.

If you plan to use GitHub, you need to make sure your local Git can pull from and push to GitHub.com. That is beyond the scope of this article, but see the [Connect to GitHub](https://happygitwithr.com/push-pull-github.html) section in Happy Git. You probably don't want to enter your username and password all the time, so either [cache credentials for HTTPS](https://happygitwithr.com/credential-caching.html) or [set up SSH keys](https://happygitwithr.com/ssh-keys.html). Then set the `usethis.protocol` option to match, i.e. to either "https" or "ssh". This setup makes functions like `usethis::use_github()` much easier to use.

If you want to use RStudio to work with Git (and therefore GitHub, see previous paragraph), you need to make sure RStudio can find your Git executable. This usually "just works". The [Connect RStudio to Git and GitHub](https://happygitwithr.com/rstudio-git-github.html) section of Happy Git helps you confirm that all is well. If all is not well, there are also [troubleshooting tips](https://happygitwithr.com/rstudio-see-git.html).

## Get and store a GitHub personal access token

A GitHub personal access token (PAT) is required if you want to use `usethis::use_github()` and `usethis::create_from_github(..., fork = TRUE)`. In both cases, usethis must create a new GitHub repository on your behalf. This is **not** a regular Git operation (like pulling and pushing to existing repos) and your usual method of providing credentials to GitHub does not work for this.

Other functions, such as the `pr_*()` family, may also use your GitHub PAT for fetch/pull/push if you are using the HTTPS protocol.

```{r, eval = FALSE}
library(usethis)
create_github_token()
```

`usethis::create_github_token()` takes you to a pre-filled form to request a PAT. You can get to the same page in the browser by clicking on "Generate new token" from <https://github.com/settings/tokens>.

As the page says, you must **store this token somewhere** because you'll never be able to see it again, once you leave that page or close the window. If you somehow goof this up, just generate a new PAT and, so you don't confuse yourself, delete the lost token.

In the moment, I usually copy the PAT to the clipboard for the next step. (I also store this PAT in my general password manager, for redundancy.)

It is customary to make this PAT available to R sessions as the value of the `GITHUB_PAT` environment variable. `create_github_token()` provides advice on how to do this, also repeated here. 

### Git credential store

This is a new approach made possible by the [credentials package](https://docs.ropensci.org/credentials/), which is used for credential handling by the [gert package](https://docs.ropensci.org/gert), which is how usethis does its Git work. The approach is described fully in the rOpenSci tech note [A better way to manage your GitHub personal access tokens](https://ropensci.org/technotes/2020/07/07/github-pat/).

`credentials::set_github_pat()` triggers a prompt where you can register your PAT with your operating system's Git credential store. That's something you do once. Or, rather, once per machine, per PAT.

From then on, executing `credentials::set_github_pat()` will cause that existing PAT to be re-validated and, assuming that succeeds, assigned to the `GITHUB_PAT` environment variable. You might put this in your `.Rprofile` startup file (remember `edit_r_profile()` helps you get this file open for editing):

```{r eval = FALSE}
# suitable for your .Rprofile
if (requireNamespace("credentials", quietly = TRUE)) {
  credentials::set_github_pat(verbose = FALSE)
}
```

However, note that usethis will soon execute this command for you when it needs the PAT.

### `.Renviron`

For years, this has been the most common way to make the PAT available. It still works, but is less secure than the approach described just above.

Here we set the `GITHUB_PAT` environment variable in the `.Renviron` file which is consulted at startup. `usethis::edit_r_environ()` will open `.Renviron` for editing.

```{r, eval = FALSE}
library(usethis)
edit_r_environ()
```

Add a line like this, **but substitute your PAT**:

```{sh, eval = FALSE}
GITHUB_PAT=8c70fd8419398999c9ac5bacf3192882193cadf2
```

Make sure this file ends in a newline! Lack of a newline can lead to silent failure to load this environment variable, which can be tricky to debug.

Restart R for this to take effect.

### Confirm your PAT is configured

Here are ways to confirm your PAT is ready to go. You can execute `usethis::github_token()`:

```{r}
library(usethis)
github_token()
```

You could inspect the `GITHUB_PAT` environment variable directly:

```{r, eval = FALSE}
Sys.getenv("GITHUB_PAT")
```

The PAT situation is also part of what's reported in `git_sitrep()`:

```{r, eval = FALSE}
git_sitrep()
```

Now usethis (and several other packages) can find this PAT automatically in order to deal with GitHub on your behalf.

## Prepare your system to build packages from source

As you participate more in R development, you will inevitably want to run development versions of other people's packages, i.e. not the version available from CRAN. A typical way to do this is to install a package from GitHub with `devtools::install_github("OWNER/REPO")`.

But, unlike using `install.packages()` and CRAN, you will be downloading and installing a *source* package, not a *binary* package. This means your system needs to be set up for building R packages. And, before long, you will need to build an R package with compiled code in it.

A full description of setting up an R development environment is beyond the scope of this article, but we give some pointers and diagnostics to get you started.

Update R and all of your packages. And expect to keep doing so frequently.

**Mac OS**: A convenient way to get the tools needed for compilation is to install Xcode Command Line Tools. Note that this is *much smaller* than full Xcode. In a shell, enter `xcode-select --install`. For installing almost anything else, consider using [Homebrew](https://brew.sh).

**Windows**: Install Rtools. This is not an R package! It is "a collection of resources for building packages for R under Microsoft Windows, or for building R itself". Go to <https://cran.r-project.org/bin/windows/Rtools/> and install as instructed.

There are functions that verify whether R package build tools are installed and available. The home and name of this function is in flux, as it is affected by the restructing of the devtools universe (written early 2018).

  * `pkgbuild::check_build_tools()` is the most current function. It can be installed via `install.packages("pkgbuild")`.
  * `devtools::has_devel()` was a function in devtools, at least up to v1.13.5, so you might still see references to that. All functionality related to package building now lives in pkgbuild. But once you have installed pkgbuild, you should just use the function `pkgbuild::check_build_tools()` instead.
