---
title: "usethis setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{usethis setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

You will get the most out of usethis if you do some setup. These setup tasks do not need to be done all at once and or even done at all. But usethis can offer the most support for package development and Git/GitHub workflows with some advance configuration. usethis can even help you with this!

Key steps that accelerate your R development workflow (details on *how* to do all this follow):

  * Make usethis available in all your interactive R sessions.
  * Install Git and configure your `user.name` and `user.email`. If you use RStudio, make sure RStudio can find your Git executable.
  * Sign up for a free GitHub.com account. Make sure you can pull/push from your local computer to GitHub.com.
  * Get a personal access token from GitHub.com and make it available in R sessions.
  * Provide usethis with info you'd like in the DESCRIPTION file of all new R packages you create.
  
As you participate more in the R development scene, you will need to be able to install packages from source that contain compiled code, e.g. C or C++. This will come up even if *you* do not write such packages (yet). We touch on this at the end.

## Use usethis in interactive work

```{r, eval = FALSE}
library(usethis)
use_usethis()
```

`usethis::use_usethis()` helps you make usethis available whenever you are working with R interactively. It opens `.Rprofile` for editing, shows the necessary code snippet in the console, and puts it on your clipboard. Paste, save, close, restart R. Done.

This is the snippet that needs to go in `.Rprofile`:

```{r, eval = FALSE}
if (interactive()) {
  suppressMessages(require(usethis))
}
```

In the long-run, devtools should make all of usethis's functions available and it should feel like they are part of devtools itself. But while the "devtools diaspora" is in process (written in first half of 2018), you may need to work with both usethis and devtools. If you attach both, we recommend attaching devtools, then usethis. In `.Rprofile`, that would like so:

```{r, eval = FALSE}
if (interactive()) {
  suppressMessages(require(devtools))
  suppressMessages(require(usethis))
}
```

You can read more about `.Rprofile` in the official [Startup documentation](https://stat.ethz.ch/R-manual/R-patched/library/base/html/Startup.html) or in Kevin Ushey's blog post [RProfile Essentials](http://kevinushey.github.io/blog/2015/02/02/rprofile-essentials/).

## Install Git, configure `user.name` and `user.email`

Please see [Happy Git and GitHub for the useR](http://happygitwithr.com/install-git.html) for instructions on how to install Git. It is beyond the scope of this article. usethis itself does not actually need the Git that you install, because it uses the [git2r](https://cran.r-project.org/package=git2r) package which wraps [libgit2](https://libgit2.github.com). But, chances are, you want to do normal Git things, like diff and commit and push, and for that you must install Git. If you want to do such things from inside RStudio, you must [make sure RStudio can find Git](http://happygitwithr.com/rstudio-git-github.html).

Once Git is installed, introduce yourself to Git.

```{r, eval = FALSE}
library(usethis)
use_git_config(user.name = "Jane Doe", user.email = "jane@example.com")

## confirm the config
use_git_config()
```

`usethis::use_git_config()` helps you configure your `user.name` and `user.email`. Substitute **your name** and **your email address**.

What user name should you give to Git? This does not have to be your GitHub username, although it can be. Another good option is your actual first name and last name. Your commits will be labelled with this name, so this should be informative to potential collaborators.

What email should you give to Git? This must be the email associated with your GitHub account.

### Git details

The code chunk above is doing the equivalent of these operations with command line git:

```{sh, eval = FALSE}
git config --global user.name 'Jane Doe'
git config --global user.email 'jane@example.com'
git config --global --list
```

### Optional: configure Git's editor

Another Git option that many people eventually configure is the editor. This will come up if you use Git from a shell. At some point, you will fail to give Git what it wants in terms of a commit message and it will kick you into an editor. This can be distressing, if it’s not your editor of choice and you don’t even know how to save and quit. You can enforce your will by executing this in a shell (not in R!):

```{sh, eval = FALSE}
git config --global core.editor emacs
```

Substitute your preferred editor for emacs here. A popular choice is nano. The default, if you don't configure `core.editor`, is usually vim.

## Get a GitHub.com account and be able to pull/push

## Get and store a GitHub personal access token

## Store your package author details as usethis options

## Prepare your system to build packages from source

